Machine Learning for Predictive Auto-Tuning

### 2011-12-12 paper/draft1.tex
Add more intro text

Current languages and software engineering practices have conspired to make it difficult
to provide high-performance libraries for general mathematical operations on GPU devices.
The dominant approach to implementing high-performance GPU codes has been auto-tuning,
which is fundamentally slow. Auto-tuning requires running a large number of
sub-compilations and representative computations to actually {\em measure} which
of several possible implementations is fastest. General-purpose libraries such
as CUBLAS and CUFFT XXX hide this overhead by auto-tuning offline and using
intimate knowledge of the mathematical problem being solved as well as the
kernel implementations to heuristically choose a good computation strategy quickly when a
user asks for e.g. a matrix product of particular arguments.
Designing and testing libraries of this form is time-consuming, difficult, and provides little
reusable infrastructure to assist in the implementation of different computations.
Libraries for stencil computations and sparse matrix operations must start from scratch.

>>>>

% 1 Introduction

% HPC Landscape: from single proc to multi-cores

Due to power consumption and heat dissipation concerns, scientific applications
have shifted from computing platforms where performance had been primarily
driven by rises in the clock frequency of a single ‘’heavy-weight’’ processor
(with complex out-of-order control and cache structures) to a platform with
ever increasing numbers of ‘’light-weight’’ cores. Interestingly, this shift is
now not only relevant to computational sciences but to the development of all
computer systems: from ubiquitous consumer-facing devices (e.g. phones) to
high-end computer farms for web-scale applications (e.g. social networks).

% HPC Landscape: large diversity of architectures, no consensus => push to flexibility => push to sw

Although the future lies in low-power multi-core hardware designs, the field
lacks meaningful consensus on exactly how the different subsystems (memory,
communication and computation) should be efficiently integrated, modelled and
programmed. These systems have exhibited varying degrees of memory hierarchy
and multi-threading complexity and, as a consequence, they have been
increasingly relying on flexible but low-level software-controlled cache
management and parallelism \cite{BerkeleyViewOfHPC} in order to better control
and understand the various trade-offs among performance, reliability, energy
efficiency, production costs, etc. This evolution has profoundly altered the
landscape of application development: programmers are now facing a wide
diversity of low-level architectural features to write high-performance
\emph{and} portable code.

% 1.1 Motivation
% Better tools: motivation (compilers/tools/libs can’t keep up, it’s too difficult)

In this rapidly evolving landscape, the construction of \emph{general}
development tools and libraries that \emph{fully utilize} system resources
remains a daunting task \cite{?}. Even within specialized architectures from
the \emph{same} vendor, like NVIDIA’s Graphics Processing Units (GPUs) and the
Compute Unified Device Architecture (CUDA) \cite{CUDA}, many developers default
to massive amounts of manual labor to optimize CUDA code to specific input
domains. In addition, hand-tuning rarely generalizes well to new hardware
generations or different input domains, and it can also be error-prone or far
from optimal. One of the reason is that kernels can produce staggeringly large
optimization spaces \cite{K. Datta et al. Stencil computation optimization and
auto-tuning on state-of-the-art multicore architectures}. The problem is
further compounded by the fact that these spaces can be highly discontinuous
\cite{RyooHwuCGO2008}, difficult to explore and holding quasi-optimal solutions
at the edge of ‘’performance cliffs’’ (in part due to hard resource
restrictions, like the register file size).

% 1.2 related work
% auto-tuning is an answer

One of the strategies for addressing these challenges is to use automatic
methods called ‘’auto-tuning’’. The literature on auto-tuning is extensive (see
surveys in \cite{XXX}. Two different approaches to auto-tuning have emerged:
model-driven optimization and empirical optimization.

The model-driven optimization approach uses analytical abstractions to model
the hardware architectures, the possible code transformations and their complex
interactions \cite{XXX}. Even though highly-accurate analytical models are
generally intractable to build, this approach has been quite successful in the
past at accelerating serial code \cite{XXX} with simplified but general
abstractions. However, large speed-ups for parallel code require more accurate
high-dimensional models and since this approach is bound by the quality and
scalability of its abstraction, it has been less suited for highly-specialized
kernels. This approach has been dominant in the compiler community \cite{XXX}.
As a result, it has generally been applied at compile-time where important
run-time characteristics such as input domains are missing. These drawbacks
render the model-driven optimization approach less attractive for
high-performance library developers.

% XXX: see gdoc and merge this below

\subsection{Case Study}
Current languages and software engineering practices have conspired to make it difficult
to provide high-performance libraries for general mathematical operations on GPU devices.
The dominant approach to implementing high-performance GPU codes has been auto-tuning,
which is fundamentally slow. Auto-tuning requires running a large number of
sub-compilations and representative computations to actually {\em measure} which
of several possible implementations is fastest. General-purpose libraries such
as CUBLAS and CUFFT XXX hide this overhead by auto-tuning offline and using
intimate knowledge of the mathematical problem being solved as well as the
kernel implementations to heuristically choose a good computation strategy quickly when a
user asks for e.g. a matrix product of particular arguments.
Designing and testing libraries of this form is time-consuming, difficult, and provides little
reusable infrastructure to assist in the implementation of different computations.
Libraries for stencil computations and sparse matrix operations must start from scratch.
